---
title: "Endogenous mechanisms and time-varying global covariates"
author: "Francisco Richter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Endogenous mechanisms and time-varying global covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 6,
    fig.height = 4
)
```

```{r}
library(amore)
```

Beyond static dyadic covariates, the rate at which relational events fire
typically depends on *what has already happened* (endogenous mechanisms such as
reciprocity) and on *when in time* the process is observed (time-varying global
factors such as weekday/weekend). `simulate_relational_events()` exposes both
through the `endogenous_stats` / `endogenous_effects` and `global_covariates` /
`global_effects` argument families. This vignette walks through a minimal
example for each, then shows how they compose.

## Endogenous reciprocity

`reciprocity_count` is a built-in endogenous statistic: the value of the
statistic at candidate dyad `(s, r)` is the number of past events `(r, s)`. A
positive coefficient means an event raises the future rate of its reverse
dyad.

```{r}
set.seed(2024)
actors <- as.character(1:10)
true_beta <- 0.6

cc <- simulate_relational_events(
  n_events = 1200,
  senders = actors,
  receivers = actors,
  baseline_rate = 1,
  allow_loops = FALSE,
  n_controls = 1,
  endogenous_stats = "reciprocity_count",
  endogenous_effects = true_beta
)
head(cc)
```

Recover the coefficient with a one-parameter conditional logit fit on the
within-stratum statistic difference:

```{r, message=FALSE, warning=FALSE}
library(mgcv)
cases    <- cc[cc$event == 1L, ]
controls <- cc[cc$event == 0L, ]
cases    <- cases[order(cases$stratum), ]
controls <- controls[order(controls$stratum), ]
fit_df <- data.frame(
  one     = 1,
  delta_r = cases$reciprocity_count - controls$reciprocity_count
)
fit <- gam(one ~ delta_r - 1, family = "binomial", data = fit_df)
unname(coef(fit)[1])
```

The estimate sits in the same ballpark as the simulated `true_beta = 0.6`.

## Time-varying global covariates

`global_covariates` is a data.frame with a strictly increasing `time_start`
column and one numeric column per global covariate. Between two breaks the
covariate value is constant. Internally, `simulate_relational_events()` uses a
boundary-aware Gillespie scheme that redraws the next waiting time whenever
the proposed event would jump into the next interval.

```{r}
set.seed(2024)
gc <- data.frame(
  time_start = seq(0, 10, by = 1),
  weekday    = rep(c(0, 1), length.out = 11)
)
ev <- simulate_relational_events(
  n_events = 200,
  senders = letters[1:5],
  receivers = letters[1:5],
  baseline_rate = 0.3,
  horizon = 11,
  global_covariates = gc,
  global_effects = c(weekday = 3)
)

share_weekday <- mean(ev$weekday == 1)
share_weekday
```

With `exp(3) ~= 20` weekday-to-weekend rate ratio, the bulk of realised events
falls in `weekday == 1` intervals.

## Composing endogenous and global

The two features can be active at the same time. The per-step total weight is
recomputed from the current endogenous state and then rescaled by the global
multiplier. The output frame carries one column per endogenous statistic and
one column per global covariate.

```{r}
set.seed(7)
actors <- letters[1:5]
gc <- data.frame(time_start = c(0, 2, 4, 6), weekday = c(1, 0, 1, 0))

ev <- simulate_relational_events(
  n_events = 60,
  senders = actors,
  receivers = actors,
  baseline_rate = 1,
  horizon = 7,
  endogenous_stats = "reciprocity_count",
  endogenous_effects = c(reciprocity_count = 0.4),
  global_covariates = gc,
  global_effects = c(weekday = 1.5)
)
head(ev)
```

Both the `reciprocity_count` column (endogenous state at event time) and the
`weekday` column (global covariate at event time) appear in the output,
ready for downstream conditional-logit or partial-likelihood inference.

## Caveat

The current endogenous-state implementation maintains a single
`(senders Ă— receivers)` reciprocity matrix and requires `senders` and
`receivers` to be the same character vector in the same order (a one-mode
network). Passing different sender/receiver sets while using
`endogenous_stats` will throw a clear error. Bipartite/two-mode support is
on the roadmap.
